---
title: "Sampling and estimation -- Demo"
author: "Jacques van Helden"
date: '`r Sys.Date()`'
output:
  html_document:
    code_folding: hide
    self_contained: no
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  slidy_presentation:
    self_contained: no
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: no
    smaller: yes
    theme: cerulean
    toc: yes
    toc_float: yes
    widescreen: yes
  word_document:
    toc: yes
    toc_depth: 3
subtitle: Probabilities and statistics for bioinformatics (STAT1)
transition: linear
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, size="huge"}
library(knitr)
## Default parameters for displaying the slides
knitr::opts_chunk$set(
  echo = TRUE, 
  eval = TRUE, 
  fig.width = 7, 
  fig.height = 5, 
  fig.align = "center", 
  fig.path = "figures/",
  size = "tiny", 
  warning = FALSE, 
  results = TRUE, 
  message = FALSE, 
  comment = "")

dir.main <- "~/stat1"
dir.current <- file.path(dir.main, "practicals", "01_intro_R")
setwd(dir.current)

```

## Generating a population

```{r parameters}
N <- 1000000
mu <- 50
sigma <- 4


```

1. Generate the observations for a population of size $N= `r N`$, following a normal distribution with mean $\mu = `r mu`$ and standard deviation $\sigma = `r sigma`$. 

```{r generate_population_values}
# help("Normal")

X <- rnorm(n = N, mean = mu, sd = sigma)


popMean <- mean(X)
popVar <- mean((X - popMean)^2)
popSD <- sqrt(popVar)

result <- data.frame( 
  "pop mean" = popMean,
  "pop var" = popVar,
  "pop SD" = popSD
)
print(result)

```

2. Draw an histogram representing the distribution of values in the population. 

```{r population_histogram, fig.cap="Distribution of numbers in the population. ", fig.height=4, fig.width=10, out.width = "80%"}
hist(x = X, 
     breaks = 100, 
     main = "Population", 
     las = 1, 
     col = "#BBDDEE", 
     xlab = "values", 
     ylab = "")
```

## Sampling

```{r sampling_parameters}
n <- 256

```

Draw a sample of size $n = `r n`$, compute its mean and standard deviation in 2 ways. 

- with the canonical formulas
- with the R functions `mean(), var(), sd()`.


```{r sampling}

# help("sample")
x <- sample(x = X, size = n, replace = FALSE)

m <- mean(x) # sample mean

s2 <-  sum((x - m)^2) / n # sample variance
s <- sqrt(s2) # sample standard dev

varFromR <- var(x) # result of the R var() function on sample values
sdFromR <- sd(x) # result of the R sd() function on sample values

result <- data.frame( 
  "sample mean" = m,
  "sample var" = s2,
  "R variance" = varFromR,
  "sample sd" = s,
  "R sd" = sdFromR
)
print(result)


# help(var)

```


```{r replicate_sampling}
repl <- 1000

sampleMeans <- replicate(
  n = repl, 
  expr = mean(sample(x = X, size = n, replace = FALSE)))
meanOfTheMeans <- mean(sampleMeans)
varOfTheMeans <- var(sampleMeans)
sdOfTheMeans <- sd(sampleMeans)

hist(sampleMeans, main = "Sample means", breaks = 25)

sampleVariancesFromR <- replicate(
  n = repl, 
  expr = var(sample(x = X, size = n, replace = FALSE)))
meanOfTheVariancesFromR <- mean(sampleVariancesFromR)


```

The mean of the sample means is `r meanOfTheMeans`. 


Replicate this sampling experiment  `r repl` times, and compute 

- the mean of the sample means
- the mean of the sample variances








