---
title: "Practical -- Microarray analysis -- Data loading and exploration"
author: "Jacques van Helden"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    highlight: zenburn
    self_contained: no
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  word_document:
    toc: yes
    toc_depth: '3'
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
font-import: http://fonts.googleapis.com/css?family=Risque
subtitle:  (STAT2)
font-family: Garamond
transition: linear
---

```{r include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)
options(width = 300)
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/02_combinatorix_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, eval = TRUE, 
  warning = FALSE, message = FALSE, 
  results = TRUE, comment = "")
# knitr::asis_output("\\footnotesize")

```


## Introduction

In this practical, we will load a dataset that will be used as study case to apply different approaches of multivariate analysis:

- data exploration
- multidimensional scaling
- differential analysis
- clustering (unsupervised classification)
- supervised classification

## Study case

**Reference:** Den Boer ML *et al.* (2009). A subtype of childhood acute lymphoblastic leukaemia with poor treatment outcome: a genome-wide classification study. Lancet Oncol. 2009 10:125-34. 

- **DOI**: [[doi: 10.1016/S1470-2045(08)70339-5](http://doi.org/10.1016/S1470-2045(08)70339-5)]
- **Pubmed**: [[PMID 19138562](https://www.ncbi.nlm.nih.gov/pubmed/19138562)]. 
- **Raw data** available at Gene Expression Omnibus, series [[GSE13425](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE13425)]
- **Preprocessed data**: <https://github.com/jvanheld/stat1/raw/master/data/DenBoer_2009/>.

## Data pre-processing

The raw microarray data has been pre-processed in order to dispose of a ready-to-use dataset. pre-processing included 

- filtering of barely detected or poorly expressed genes, 
- log2 transformation to normalise the raw measurements
- between-sample standardisation to enable comparison between the different samples.

## Availability of the pre-processed data


The preprocessed data is available here:  <https://github.com/jvanheld/stat1/raw/master/data/DenBoer_2009/>.

It contains the following files.

| File | Contents | Structure |
|--|--|--|
| [GSE13425_group_descriptions.tsv.gz](../../data/DenBoer_2009/GSE13425_group_descriptions.tsv.gz) | Description of the patient groups | Tab-delimited file with one row per group and one column per type of description (group name, label) |
| [phenoData_GSE13425.tsv.gz](../../data/DenBoer_2009/phenoData_GSE13425.tsv.gz) | Metadata (sample descriptions) | Tab-delimited file with one row per biological sample (one per patient) and one column per type of information about the biological sample |
| [GSE13425_Norm_Whole.tsv.gz](../../data/DenBoer_2009/GSE13425_Norm_Whole.tsv.gz) | Normalised microarray data | Tab-delimited file with one row per gene and one column per patient |
| [GSE13425_AMP_Whole.tsv.gz](../../data/DenBoer_2009/GSE13425_AMP_Whole.tsv.gz) | Detection status of each gene in each sample (Absent, Marginal, Present) | Tab-delimited file with one row per gene and one column per patient|

## Data download

Write an R script that perform the following operations

1. Creates a directory to store a local copy for this practical on your computer. (`~/STAT2_CMB_practicals/den-boer-2009/`).
2. Creates a sub-directory for the data  (`~/STAT2_CMB_practicals/den-boer-2009/data/`).
3. Lists the files available on the data Web site (<https://github.com/jvanheld/stat1/raw/master/data/DenBoer_2009/>).
4. For each of these files, checkS if it is already present in your local local data directory, and  downloads it if it is not the case. 

Try to make your code re-usable, in the perspective to apply it soon in order to download data sets from other web sites. 

```{r}
#' @title Download a set of files from a Web site
#' @author Jacques van Helden
#' @description Download a set of files from a Web site and store them in a destination directory if they do not exist yet there. 
#' @param files a vector with the file names
#' @param dataURL the URL of the folder containing the files to download
#' @param destDir local destination directory. Will be created if it does not exist. 
#' @return no return value
#' @export
DownloadMyFiles <- function(files, 
                            dataURL,
                            destDir) {
  
  ## Create local directory
  dir.create(destDir, recursive = TRUE, showWarnings = FALSE)
  
  for (f in files) {
    
    ## Destination file
    destFile <- file.path(destDir, f)
    
    ## Check if file exists
    if (file.exists(destFile)) {
      message("skipping download because file exists: ", destFile)
    } else {
      sourceURL <- file.path(dataURL, f)
      download.file(url = sourceURL, destfile = destFile)
    }
  }
}

```


```{r}
## URL fo the folder containing the data fle
dataURL <- "https://github.com/jvanheld/stat1/raw/master/data/DenBoer_2009/"

files <- c(
  expr = "GSE13425_Norm_Whole.tsv.gz",
  pheno = "phenoData_GSE13425.tsv.gz",
  groups = "GSE13425_group_descriptions.tsv.gz"
)

## Local data dir
destDir <- "~/STAT2_CMB_practicals/den-boer-2009/data/"

## Call the function to download the files
DownloadMyFiles(files = files, dataURL = dataURL, destDir = destDir)

kable(data.frame(list.files(destDir)), caption = "Content of the destination directory after file download. ")

```

## Data loading

Write an R script that loads the data tables from your local data directory. 

```{r}

exprTable <- read.table(file.path(destDir, files["expr"]))
dim(exprTable)

```


## Marginal statistics

Write an R script that computes marginal statistics (mean, sd, min, percentiles 5, 10, 25, 50, 75, 90, 95, max, IQR) on each row of the normalised data table (one row corresponds to one gene). 

Do the same for each column (patient). 

## Empirical distributions

Draw a plot that displays the empirical distribution of normalised expression values in the whole data table.

Draw another plot that displays the empirical distribution of normalised expression values in the different samples (one polygon frequency per sample).

Compute a table with the mean expression profile per cancer type (one row per gene, one column per cancer type) and draw them with box plots. 

## Sample classes

Load the pheno table. We will use the three following columns: 

- `Sample.GEO.ID`: the identifier of the sample in the Gene Expression Omnibus (GEO) database.
- `Sample.title`:  cancer type of each sample. 
- `sample.labels`: A short label (1 or 2 letters) for each cancer type

Cound the number of samples per group and draw a barplot with the result.


